### 2019-12-06 Manuscript/manuscript.tex
small edits to manuscript text including adding mention of LEH2013 model

Laurentia refers to the craton that forms the Precambrian core of North America (Fig. \ref{fig:Laurentia_map}). Laurentia is comprised of multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}. A period of accretionary and collision orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca. 1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}. As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks was not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016b}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade tectonism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between Wyoming and Superior. Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast. Overall, the collision of these Archean microcontinents between ca. 1.9 and 1.8 Ga lead to rapid amalgamation of the majority of the Laurentia craton.

@@

In the paleogeographic model framework of \cite{Pehrsson2015a}, the collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previous been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}). The \cite{Pehrsson2015a} model posits that this period terminal collision not only resulted in the amalgamation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.

Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny. Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65–1.60 Ga Mazatzal Orogeny and associated plutonism lead to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}. Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion. An interval of major plutonism occurred ca. 1.48–1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}. This plutonism is likely due to crustal melting within a back-arc region of ca. 1.50 to 1.43 Ga accretionary orogenesis \citep{Bickford2015a}. Younger magmatic activity ca. 1.37 Ga  of the Southern Granite–Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}. While an active margin interpretation with magmatism in back‐arc setting has gained traction within the literature with additional data, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).

@@

Accretionary orogenesis continued along the (south)east margin of Laurentia with the arc-continent collision of the ca. 1.25-1.22 Ga Elzevirian orogeny \citep{McLelland2013a}. The subsequent ca. 1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a terrane comprised of amalgamated arc volcanics and associated metasediments and is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collision orogenesis of the Grenvillian orogeny \citep{McLelland2010a}. In the latest Mesoproterozoic (ca. 1.11-1.08 Ga), a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and Paleoproterozoic provinces. This Midcontinent Rift lead to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.  Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}. The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism through the orogen \citep{McLelland2010a}. The orogeny is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a}.

@@

This late Paleoproterozoic and Mesoproterozoic tectonic history provides significant constraints on paleogeographic reconstructions. In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from the initiation of the Yavapai orogeny (ca. 1.71 Ga) to the end of the Shawinigan orogeny (ca. 1.06 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}. This constraint is incorporated into models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g. Amazonia in \citealp{Elming2009a}). The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself requires a collision between Laurentia and (an)other continent(s) ca. 1080 Ma -- the geological observation of which first lead to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}. Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas, up through the Blue Ridge Appalachian inliers, through Ontario and up to the Labrador Sea. This extensive and major collision orogenic history recorded in Laurentia remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition. The term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt. The timeline of orogenesis on the Laurentia margin has more nuanced constraints that this usage and these constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig. \ref{fig:tectonic_history}).

@@

Extensive rifting that was followed by thermal subsidence occurred along the southeast to east Laurentia margin leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of modern-day North Carolina and Virginia. Rifting ca. 620 to 580 Ma initiates in the region from modern-day New York to Newfoundland and by ca. 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin. The last phases of this rifting appears to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}. Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated associated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}. As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting of a major craton as the record that lingers on the craton is similar. One interpretation is that there was successful break-up along the eastern margin during the ca. 580 to 550 Ma interval of rifting prior to the ca. 539 Oklahoma aulacogen rifting that liberated the Cuyania microcontinent. The Maz–Arequipa–Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting. Orogenesis between the MARA block and the Rio de Plata and Kalahari in the ca. 530 Ma Pampean orogeny \citep{Casquet2018a} predated the collision of Cuyania during the ca. 460 Ma Famatinian orogeny in West Gondwana \citep{Rapalini2018a}.

@@

Even without considering other continents on Earth, the geological record of Paleoproterozoic collisional of Archean provinces combined with accretionary orogenesis at that time and through the rest of the Paleoproterozoic and Mesoproterozoic Eras provides very strong evidence for mobile plate tectonics driving Laurentia's evolution throughout the past 2 billion years. This tectonic history inferred from geological data can be enhanced through integration with the paleomagnetic record.

@@

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results (Evans et al., this volume). While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland. An `A' rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. A `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}. Many of these poles are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of such poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained including its age. For example, there are rich data associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions. However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are well-constrained (Table 2). Additional not-rated poles included in the present compilation is the new pole for the ca. 1144 Ma Ontario lamprophyre dykes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group as presented in \cite{Eyster2019a} are also included.

@@

There are poles in the compilation for the Slave, Wyoming, Rae, Superior and North Atlantic provinces prior to Laurentia amalgamation (Fig. \ref{fig:Laurentia_poles} and Table 2). Overall, these data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly. A lingering question raised in \citet{Hoffman1988a} that still remains is to what extent the Archean provinces each had independent drift histories with significant separation or shared histories before experiencing fragmentation and reamalgamation. The strongest analysis in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}. High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their modern-day relative orientation to one another either and having distinct pole paths as constrained by 5 time periods of nearly coeval poles from 2.23 and 1.89 Ga (Fig. \ref{fig:Superior_Slave_recons}; \citealp{Buchan2016a}. These data provide paleomagnetic support for the Superior and Slave provinces having independent histories of differential motion. They also support the hypothesis that the Trans-Hudson orogeny is the result of terminal collision associated with the closure of an ocean basin between the Superior province and the Hearne+Rae+Slave provinces. Reconstructions developed for this chapter of the Superior and Slave provinces using these poles are shown in Figure \ref{fig:Superior_Slave_recons} and illustrate the difference in implied orientation and paleolatitude that results from these well-constrained poles.

@@

Another challenging portion of the Laurentia record is that for the Ediacaran Period where there is little consistency between poles of similar age (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitudes}). As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 615 and 565 Ma. One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran that has influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}. Another possibility is that the lack of congruency between poles in this point in the record is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Bono2019a}. Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the composite (Fig. \ref{fig:Laurentia_paleolatitude}).

>>>>

Laurentia refers to the craton that forms the Precambrian core of North America (Fig. \ref{fig:Laurentia_map}). Laurentia is comprised of multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}. A period of accretionary and collision orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca. 1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}. As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks were not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016b}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade tectonism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between Wyoming and Superior. Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast. Overall, the collision of these Archean microcontinents between ca. 1.9 and 1.8 Ga lead to rapid amalgamation of the majority of the Laurentia craton.

@@

In the paleogeographic model framework of \cite{Pehrsson2015a}, the collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}). The \cite{Pehrsson2015a} model posits that this period terminal collision not only resulted in the amalgamation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.

Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny. Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism lead to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}. Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion. An interval of major plutonism occurred ca. 1.48 to 1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}. This plutonism is likely due to crustal melting within a back-arc region of ca. 1.50 to 1.43 Ga accretionary orogenesis \citep{Bickford2015a}. Younger magmatic activity ca. 1.37 Ga of the Southern Granite-Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}. While an active margin interpretation with magmatism in back-arc setting has gained traction within the literature with additional data, the tectonic setting is often described as enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).

@@

Accretionary orogenesis continued along the (south)east margin of Laurentia with the arc-continent collision of the ca. 1.25 to 1.22 Ga Elzevirian orogeny \citep{McLelland2013a}. The subsequent ca. 1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a terrane comprised of amalgamated arc volcanics and associated metasediments and is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collision orogenesis of the Grenvillian orogeny \citep{McLelland2010a}. In the latest Mesoproterozoic (ca. 1.11 to 1.08 Ga), a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and Paleoproterozoic provinces \citep{Cannon1992b}. This Midcontinent Rift led to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.  Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}. The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism through the orogen \citep{McLelland2010a}. The orogeny is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a}.

@@

This late Paleoproterozoic and Mesoproterozoic tectonic history provides significant constraints on paleogeographic reconstructions. In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from the initiation of the Yavapai orogeny (ca. 1.71 Ga) to the end of the Shawinigan orogeny (ca. 1.06 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}. This constraint is incorporated into models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g. Amazonia in \citealp{Elming2009a}). The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself requires a collision between Laurentia and (an)other continent(s) ca. 1080 Ma -- the geological observation of which first lead to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}. Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas, up through the Blue Ridge Appalachian inliers, through Ontario and up to the Labrador Sea. This extensive and major collision orogenic history recorded in Laurentia remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition. The term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt. The timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage and these constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig. \ref{fig:tectonic_history}).

@@

Extensive rifting that was followed by thermal subsidence occurred along the southeast to east Laurentia margin leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of modern-day North Carolina and Virginia. Rifting ca. 620 to 580 Ma initiates in the region from modern-day New York to Newfoundland and by ca. 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin. The last phases of this rifting appears to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}. Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated associated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}. As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting of a major craton as the record that lingers on the craton is similar. One interpretation is that there was successful break-up along the eastern margin during the ca. 580 to 550 Ma interval of rifting prior to the ca. 539 Oklahoma aulacogen rifting that liberated the Cuyania microcontinent. The Maz-Arequipa-Rio Apa (MARA) block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting. Orogenesis between the MARA block and the Rio de Plata and Kalahari in the ca. 530 Ma Pampean orogeny \citep{Casquet2018a} predated the collision of Cuyania during the ca. 460 Ma Famatinian orogeny in West Gondwana \citep{Rapalini2018a}.

@@

Even without considering other continents on Earth, the geological record of Paleoproterozoic collisional of Archean provinces combined with accretionary orogenesis at that time and through the rest of the Paleoproterozoic and Mesoproterozoic Eras provides strong evidence for mobile plate tectonics driving Laurentia's evolution throughout the past 2 billion years. This tectonic history inferred from geological data can be enhanced through integration with the paleomagnetic record.

@@

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results (Evans et al., this volume). While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland. An `A' rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. A `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}. Many of these poles in the Paleomagia database are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of such poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained including its age. For example, there are rich data associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions. However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are well-constrained (Table 2). Additional not-rated poles included in the present compilation is the new pole for the ca. 1144 Ma Ontario lamprophyre dykes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group as presented in \cite{Eyster2019a} are also included.

@@

There are poles in the compilation for the Slave, Wyoming, Rae, Superior and North Atlantic provinces prior to Laurentia amalgamation (Fig. \ref{fig:Laurentia_poles} and Table 2). Overall, these data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly. A lingering question raised in \citet{Hoffman1988a} that still remains is to what extent the Archean provinces each had independent drift histories with significant separation or shared histories before experiencing fragmentation and reamalgamation. The strongest analysis in this regard comes from comparisons between paleomagnetic poles between the Superior and Slave provinces \citep{Buchan2009a, Mitchell2014a, Buchan2016a}. High-quality paleomagnetic poles from these two provinces provide strong support for differential motion between the Superior and Slave provinces between 2.2 and 1.8 Ga with the two provinces not being in their modern-day relative orientation to one another and having distinct pole paths as constrained by 5 time periods of nearly coeval poles from 2.23 and 1.89 Ga (Fig. \ref{fig:Superior_Slave_recons}; \citealp{Buchan2016a}. These data provide paleomagnetic support for the Superior and Slave provinces having independent histories of differential motion. The data also support the hypothesis that the Trans-Hudson orogeny is the result of terminal collision associated with the closure of an ocean basin between the Superior province and the Hearne+Rae+Slave provinces. Reconstructions developed for this chapter of the Superior and Slave provinces using these poles are shown in Figure \ref{fig:Superior_Slave_recons} and illustrate the difference in implied orientation and paleolatitude that results from these well-constrained poles.

@@

Another challenging portion of the Laurentia record is that for the Ediacaran Period where there is little consistency between poles of similar age (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 615 and 565 Ma. One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran that has influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}. Another possibility is that the lack of congruency between poles in this point in the record is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Bono2019a}. Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the composite (Fig. \ref{fig:Laurentia_paleolatitude}).

%A nice touch would be to add Paleoproterozoic, Meso, Neo and Phanero onto this figure
\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
\caption{\small{\textbf{Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79\textdegree N, lon=92.10\textdegree W). The paleomagnetic poles are compiled in Table 2. Bottom panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}.}}}
\label{fig:Laurentia_paleolatitude}
\end{figure}
